---
title: "Advertising Simulator"
author: "Günter J. Hitsch"
date: "January 24, 2017"
output: pdf_document
urlcolor: blue
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
                      fig.width = 4.5, fig.height = 3, fig.align = "right")
```

```{r}
library(data.table)
library(ggplot2)
```




#### Model

Demand function:
$$\log(Q_t) = \alpha + \gamma g_t$$
Goodwill:
$$g_{t} = \log(a_{t}) + \delta \log(a_{t-1}) + \dots + \delta^{L} \log(a_{t-L})$$


#### Sales response simulation function

- `T`: prediction horizon
- `adv_base_level`: Constant base level of advertising
- `adv_insert`: Vector of advertising levels that are added to the base level (values can be negative)
- `insert_t`: Period when first element in `adv_insert` is added to the base level of advertising (between 1 and `T`)
- `param`: List of model parameters

```{r}
advertisingResponse <- function(T, adv_base_level, adv_insert, insert_t, param) {
   if (insert_t > T) stop("Insert time outside simulation period")
   
   geom_weights = cumprod(c(1.0, rep(param$delta, param$L)))
   geom_weights = sort(geom_weights)
   
   # Pad with zero values before and after insertin period
   adv_insert = c(rep(0, insert_t-1), adv_insert, rep(0, T))
   
   log_Q = rep(NA, times = T)
   adv   = rep(adv_base_level, param$L)
   for (t in 1:T) {
      adv = c(adv, adv_base_level + adv_insert[t])
      g   = sum(geom_weights*tail(log(adv), param$L+1))
      log_Q[t] = param$alpha + param$gamma*g
   }
   
   L = length(adv)
   DT = data.table(time  = 1:T,
                   adv   = adv[(L-T+1):L],
                   log_Q = log_Q)
   
   return(DT)
}
```


#### Parameter settings

```{r}
parameters = list(alpha = 1.0,
                  gamma = 1.0,
                  delta = 0.8,
                  L     = 10   )
```


#### Simulation


```{r}
T              = 20
adv_base_level = 1
insert_t       = 4
adv_insert     = c(4, 4, 2)

DT = advertisingResponse(T, adv_base_level, adv_insert, insert_t, parameters)
```


\newpage

#### Graph the response

```{r}
ggplot(DT, aes(x = time, y = adv)) +
   geom_vline(xintercept = c(insert_t, min(insert_t+length(adv_insert), T)),
              color = "lightblue1", size = 2, alpha = 0.8) + 
   geom_step(color = "gray75", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "hotpink", size = 1.75, stroke = 0.5) +
   scale_x_continuous(minor_breaks = 1:T) +
   theme_bw()

# ggsave("Simulated-Adv.pdf", width = 6, height = 2)
```

```{r}
ggplot(DT, aes(x = time, y = log_Q)) +
   geom_vline(xintercept = c(insert_t, min(insert_t+length(adv_insert)+parameters$L, T)),
              color = "lightblue1", size = 2, alpha = 0.8) + 
   geom_step(color = "gray75", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "midnightblue", size = 1.75, stroke = 0.5) +
   scale_x_continuous(minor_breaks = 1:T) +
   theme_bw()

# ggsave("Simulated-Sales.pdf", width = 6, height = 2)
```


