---
title: "Causal Forest"
author: "Günter J. Hitsch"
date: "February 19, 2017"
output: pdf_document
urlcolor: blue
graphics: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
                      fig.width = 4.5, fig.height = 3.25, fig.align = "right")
```




## Installation

The `causalTree` package is actively being developed and it needs to be installed and compiled from sources. This is not hard, but you first need to install the necessary development environment.

#### Step 1

**Windows**: Download and install [Rtools](https://cran.r-project.org/bin/windows/Rtools/).

**Mac OS**: Download `Xcode` from the App Store. Warning: `Xcode` may take more than one hour to download. Then launch the `Xcode` application and accept the license terms. Open the `Terminal` (command line) application (installed on any Mac and located in `/Applications/Utilities`). In the `Terminal` window type:

```{}
xcode-select --install
```

An alert box should appear; click *Install* to proceed.

#### Step 2

Once step 1 is completed reboot your computer (just to be on the safe side). Then install the R package `devtools`.

\bigskip

Now you can install and build the `causalTree` package:

```{r, eval = FALSE}
library(devtools)
install_github("walterwzhang/causalTree")
```

\newpage




## Causal forest example

```{r}
library(data.table)
library(ggplot2)
library(causalTree)
library(broom)
library(knitr)
```

\bigskip


We simulate a data set that includes the following variables:

- `target` is the *treatment*, a dummy variable indicating if a customer was targeted (e-mail/catalog). The treatment assignment is random.
- `spend` is observed dollar spending.
- `recency` is the customer recency status (in months), ranging from 1 to 18.
- `web_buyer` is a dummy variable that indicates if a customer is a frequent user of the company's website.

The purchase probability $p$ takes the following form:

1. **Customers who are not targeted**. For `recency` between 1 and 6, $p=0$. Then, for `recency` between 7 and 12, $p$ increases in `recency` and takes the value $p = 0.03 \cdot (\texttt{recency} - 6)$. For all values of `recency` above 12, $p = 0.03 \cdot 6 = 0.18$.

2. **Customers who are targeted**

    (a) If `web_buyer`, then $p$ is 1.25 times the baseline purchase probability.
    (b) If not `web_buyer`, then $p$ is twice the the baseline purchase probability.
    
Spending *conditional* on a purchase is uniformly distributed on the values $80,81,\dots,119,120$, with a corresponding mean of 100. Hence, expected spending is $100 \cdot p$.

In this model, the treatment effect is non-linear in `recency`, and the treatment effect is larger (for `recency` above 6) for customers who are not a `web_buyer`.

\medskip

```{r, results = "hide"}
set.seed(941)

n_obs  = 100000        # Training
n_pred = 100000        # Prediction
n = n_obs + n_pred

customer_DT = data.table(recency   = sample.int(18, size = n, replace = TRUE),
                         web_buyer = rbinom(n, 1, 1/3),
                         target    = rbinom(n, 1, 0.5))

# Define the purchase probability p
customer_DT[recency <= 6,                 p := 0.0]
customer_DT[recency >  6 & recency <= 12, p := 0.03*(recency - 6)]
customer_DT[recency > 12,                 p := 0.03*6]
customer_DT[web_buyer == 1 & target == 1, p := 1.25*p]
customer_DT[web_buyer == 0 & target == 1, p := 2*p]

# Simulate spending data
customer_DT[, purchase   := runif(n) <= p]
customer_DT[, cond_spend := 79 + sample.int(41, size = n, replace = TRUE)]
customer_DT[, spend      := purchase*cond_spend]

training_DT = customer_DT[1:n_obs]
pred_DT     = customer_DT[(n_obs+1):n]
```

\bigskip

Now we estimate the causal forest. Note that we need to specify the treatment variable.

Note: When first estimating a causal forest, I recommend to set the `num.trees` (number of trees) option to a small value, maybe 10, to get a sense how much computation time is involved.

The `verbose` option was added for your convenience but is not part of the original package. Set to `FALSE` if you don't want to see the output messages indicating the progress in growing the random forest.

```{r, results = "hide"}
fit = causalForest(spend ~ recency + web_buyer,
                   treatment = training_DT$target,
                   data = training_DT,
                   num.trees = 1000,
                   verbose = TRUE)
```

Predict spending in the prediction sample.

```{r, results = "hide"}
pred_DT[, pred_treatment_effect := predict(fit, pred_DT)]
```

Now we create a table with the true, observed, and predicted conditional average treatment effects ($\tau$) for all values of `web_buyer` and `recency`.

```{r}
summary_DT = pred_DT[, .(tau      = 100*(mean(p[target==1]) - mean(p[target==0])),
                         tau_obs  = mean(spend[target==1]) - mean(spend[target==0]),
                         tau_pred = mean(pred_treatment_effect)),
                     keyby = .(web_buyer, recency)]
```

\newpage

```{r}
kable(summary_DT[web_buyer == 0], digits = 2)
```

```{r}
kable(summary_DT[web_buyer == 1], digits = 2)
```

\newpage


#### Model fit: causal forest

```{r}
ggplot(summary_DT[web_buyer == 0], aes(x = recency, y = tau)) +
   geom_line(color = "navyblue", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "navyblue", size = 2, stroke = 0.5) +
   geom_line(aes(x = recency, y = tau_pred),
             color = "mediumvioletred", size = 0.5) +
   geom_point(aes(x = recency, y = tau_pred),
              shape = 21, color = "gray30", fill = "mediumvioletred", size = 2, stroke = 0.5) +
   
   theme_bw()
```

```{r}
ggplot(summary_DT[web_buyer == 1], aes(x = recency, y = tau)) +
   geom_line(color = "navyblue", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "navyblue", size = 2, stroke = 0.5) +
   geom_line(aes(x = recency, y = tau_pred),
             color = "mediumvioletred", size = 0.5) +
   geom_point(aes(x = recency, y = tau_pred),
              shape = 21, color = "gray30", fill = "mediumvioletred", size = 2, stroke = 0.5) +
   
   theme_bw()
```

\newpage


#### Model fit: OLS

```{r, results = "hide"}
minimal_DT = training_DT[, .(spend, recency, web_buyer, target)]

fit_OLS = lm(spend ~ . + .*target + web_buyer:recency*target,
             data = minimal_DT)

pred_DT[, pred_spend_OLS := predict(fit_OLS, pred_DT)]

summary_OLS_DT = pred_DT[, .(tau_pred_OLS = mean(pred_spend_OLS[target==1])
                                          - mean(pred_spend_OLS[target==0])),
                         keyby = .(web_buyer, recency)]
summary_OLS_DT = merge(summary_OLS_DT, summary_DT[, .(web_buyer, recency, tau)],
                       by = c("web_buyer", "recency"))
```

```{r}
ggplot(summary_OLS_DT[web_buyer == 0], aes(x = recency, y = tau)) +
   geom_line(color = "navyblue", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "navyblue", size = 2, stroke = 0.5) +
   geom_line(aes(x = recency, y = tau_pred_OLS),
             color = "mediumvioletred", size = 0.5) +
   geom_point(aes(x = recency, y = tau_pred_OLS),
              shape = 21, color = "gray30", fill = "mediumvioletred", size = 2, stroke = 0.5) +
   
   theme_bw()
```

```{r}
ggplot(summary_OLS_DT[web_buyer == 1], aes(x = recency, y = tau)) +
   geom_line(color = "navyblue", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "navyblue", size = 2, stroke = 0.5) +
   geom_line(aes(x = recency, y = tau_pred_OLS),
             color = "mediumvioletred", size = 0.5) +
   geom_point(aes(x = recency, y = tau_pred_OLS),
              shape = 21, color = "gray30", fill = "mediumvioletred", size = 2, stroke = 0.5) +
   
   theme_bw()
```

\newpage


#### Model fit: OLS with polynomials

```{r, results = "hide"}
minimal_DT = training_DT[, .(spend, recency, web_buyer, target)]

fit_OLS = lm(spend ~ target*web_buyer + poly(recency,3)*web_buyer + poly(recency,3):target*web_buyer,
             data = minimal_DT)

pred_DT[, pred_spend_OLS := predict(fit_OLS, pred_DT)]

summary_OLS_DT = pred_DT[, .(tau_pred_OLS = mean(pred_spend_OLS[target==1])
                                          - mean(pred_spend_OLS[target==0])),
                         keyby = .(web_buyer, recency)]
summary_OLS_DT = merge(summary_OLS_DT, summary_DT[, .(web_buyer, recency, tau)],
                       by = c("web_buyer", "recency"))
```


```{r}
ggplot(summary_OLS_DT[web_buyer == 0], aes(x = recency, y = tau)) +
   geom_line(color = "navyblue", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "navyblue", size = 2, stroke = 0.5) +
   geom_line(aes(x = recency, y = tau_pred_OLS),
             color = "mediumvioletred", size = 0.5) +
   geom_point(aes(x = recency, y = tau_pred_OLS),
              shape = 21, color = "gray30", fill = "mediumvioletred", size = 2, stroke = 0.5) +
   
   theme_bw()
```

```{r}
ggplot(summary_OLS_DT[web_buyer == 1], aes(x = recency, y = tau)) +
   geom_line(color = "navyblue", size = 0.5) +
   geom_point(shape = 21, color = "gray30", fill = "navyblue", size = 2, stroke = 0.5) +
   geom_line(aes(x = recency, y = tau_pred_OLS),
             color = "mediumvioletred", size = 0.5) +
   geom_point(aes(x = recency, y = tau_pred_OLS),
              shape = 21, color = "gray30", fill = "mediumvioletred", size = 2, stroke = 0.5) +
   
   theme_bw()
```

```{r}
out = tidy(summary(fit_OLS))
kable(out, digits = 2)
```



