---
title: "Regression and Classification Trees"
author: "Günter J. Hitsch"
date: "February 21, 2017"
output: pdf_document
urlcolor: blue
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
                      fig.width = 7, fig.height = 6, fig.align = "right")
```

```{r, include = FALSE}
library(data.table)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(knitr)
```




## Regression trees

As an example, we use household-level private label share data aggregated at the *year* level.

```{r}
load("./Data/PL_shares_annual.RData")
```

Convert all categorical variables to factors. The `rpart` package will then understand that these variables are categorical.

```{r, results = "hide"}
categorical_vars = c("unemployed", "has_children", "female_head", "size")
PL_shares_DT[, (categorical_vars) := lapply(.SD, as.factor), .SDcols = categorical_vars]
```

Now build a regression tree (`method = "anova"`).

```{r}
tree = rpart(PL_share ~ ., data = PL_shares_DT, method = "anova",
             control = rpart.control(minsplit = 20, cp = 0.001, maxdepth = 5))
```

The cost-complexity parameter `cp` is key: If the algorithm does not split, make the parameter smaller. `minsplit` determines the minimum number of observations in a node such that a split is attempted, `maxdepth` determines the depth (length) of a tree.

```{r}
N_terminal_nodes = length(unique(tree$where))
rpart.plot(tree, tweak = 0.8, digits = 3)
```



#### Cost-complexity pruning

Grow a large tree:

```{r}
tree = rpart(PL_share ~ ., data = PL_shares_DT, method = "anova",
             control = rpart.control(minsplit = 20, cp = 0.00005))

N_terminal_nodes = length(unique(tree$where))
# rpart.plot(tree, tweak = 0.8, digits = 3)
```

Visualize the cross-validation results:

```{r}
plotcp(tree)
```

Find the optimal complexity parameter:

```{r}
index_optimal = which.min(tree$cptable[, "xerror"])
cp_optimal    = tree$cptable[index_optimal, "CP"]
size_optimal  = tree$cptable[index_optimal, "nsplit"] + 1
```

Prune the tree:

```{r}
pruned_tree = prune(tree, cp = cp_optimal)
# rpart.plot(pruned_tree)
```



\newpage

## Classification trees

Use the RFM data to predict if a customer buys.

```{r}
load("./Data/RFM-Data.RData")
```

Use the `method = "class"` option for classification.

```{r}
tree = rpart(buyer ~ ., data = rfm_DT, method = "class",
             control = rpart.control(minsplit = 20, cp = 0.00005))

rpart.plot(tree, tweak = 0.8, digits = 3)
```



